Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 9 09:38:49 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/dc55a82ecaa6c7579c80e824377410a6b61dc53f}{dc55a82ecaa6c7579c80e824377410a6b61dc53f}}

### 2017-02-27 sftrends_manu.Rnw
fig 1 with whole bay

\begin{abstract}
\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.  We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}. Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) effects of biological invasion on chlorophyll dynamics in Suisun Bay.  Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
\end{abstract}
\acresetall

@@

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}. Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years. Indicators of eutrophication can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and on characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into eutrophication dynamics in aquatic systems.

The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.  Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.  Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.  The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales. The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \cite{Jassby00,Jassby02,Jassby08}. A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \cite{Jabusch16}.  Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.  To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.

The goal of this study was to provide a comprehensive description of nutrient trends in the Delta to inform understanding of eutrophication dynamics and potential causes of water quality change in the larger Bay. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations in the Delta to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.

@@

The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest bay on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land in the Central Valley.  Water enters the Bay through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system is divided into several sub-bays, including Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.  Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.  The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.  Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.

>>>>

\begin{abstract}
\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.  We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}. Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
\end{abstract}
\acresetall

@@

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}. Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of ecosystem repsonse indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters. Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.

The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.  Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.  Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.  The unique ecological and social context of the region provides a valuable opportunity to gain insight into ecosystem properties that define water quality dynamics at different scales. In particular, the Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \ac{sfe} \cite{Jassby00,Jassby02,Jassby08}. A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.  Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.  To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.

The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations as additional case studies o demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.

@@

The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.  Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.  Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.  The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.  Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.

### 2017-02-27 sftrends_manu.tex
fig 1 with whole bay

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 2 16:00:40 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/4afb16203c36788d9f1a8c7b6ee43ed973d94495}{4afb16203c36788d9f1a8c7b6ee43ed973d94495}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

\begin{abstract}
\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.  We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}. Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) effects of biological invasion on chlorophyll dynamics in Suisun Bay.  Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
\end{abstract}
\acresetall

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 9 09:38:49 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/dc55a82ecaa6c7579c80e824377410a6b61dc53f}{dc55a82ecaa6c7579c80e824377410a6b61dc53f}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

\begin{abstract}
\noindent Recent methods for trend analysis have been developed that leverage the descriptive potential of multi-decadal monitoring data.  We apply an estuarine adaptation of the \ac{wrtds} model to describe water quality trends over four decades in the Delta region of the \ac{sfe}. Results from multiple stations in the Delta provided novel descriptions of historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with \ac{wrtds} flow-normalized data demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further demonstrate use of \ac{wrtds} to provide insight into mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  Overall, this analysis provides an ecological and management-based understanding of historical trends in the Delta as a means to interpret potential impacts of recent changes and expected trends.
\end{abstract}
\acresetall
